Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2017-05-01 pkg/glmertree/paper/treatment_interactions.Rnw
improvements in methods section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1295 255f4249-2263-4417-b1ea-8498779e27a7

We will use an artificial motivating dataset with treatment-subgroup interactions to introduce the GLMM tree algorithm. The R code used to generate the example dataset is provided in the Appendix.

The dataset consists of a set of observations on $N = 150$ patients, who were randomly assigned to one of two treatment alternatives (Treatment~1 or Treatment~2). The treatment outcome is represented by the variable depression, quantifying post-treatment depressive symptomatology. The potential moderator variables are duration, age and anxiety. Duration reflects the number of months the patient has been suffering from depression prior to treatment, age reflects patients' age in years at the start of treatment and anxiety reflects patients' total scores on an anxiety inventory administered before treatment. Finally, each patient was part of one of ten clusters, each having a different value for the random intercept, uncorrelated to the partitioning variables.

The outcome variable was generated such that there were three subgroups with differential treatment effectiveness, corresponding to the terminal nodes in Figure~\ref{fig:exampletree}: The first subgroup of patients corresponds to node~3 and shows lower post-treatment depression in Treatment~1 than in Treatment~2. The second subgroup of patients corresponds to node~4 and shows about equal post-treatment depression in both treatment conditions. The third subgroup of patients corresponds to node~5 and shows lower post-treatment depression in Treatment~2 than in Treatment~1. Thus, duration and anxiety are true partitioning or moderator variables, whereas age is not. Anticipating the final results of our analyses, the treatment-subgroup interactions are depicted in Figure~\ref{fig:example_glimmertree}, which shows the GLMM tree that accurately recovered the treatment-subgroup interactions.

@@

The rationale behind MOB is that a global parametric model may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find better-fitting models in each cell of the partition. For example, to assess the effect of treatment, we may fit a global GLM using only the treatment indicator as a predictor of the outcome. The expectation $\mu_i$ of outcome $y_i$ given the treatment regressor $x_i$ is then modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
	\label{eq:expected_value}
	E[y_i | x_i] & = & \mu_i, \\
	\label{eq:fixedeffects}
	g(\mu_{i}) & = & x_{i}^{\top}\beta,
\end{eqnarray}
%
where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function. $\beta$ is a vector of fixed-effects regression coefficients. For simplicity, in the current paper we assume only two treatment groups and no further covariates in the GLM, so $x_{i}$ and $\beta$ each have length 2. The first element of $\beta$ then corresponds to the mean of the linear predictor in the first treatment group and the second element corresponds to the mean difference in the linear predictor between the first and second treatment groups. For the continuous response variable in the motivating data set, we employ the identity link function and assume a normal distribution for the error (denoted by $\epsilon_i = y_i - \mu_i$) with mean zero and variance $\sigma_{\epsilon}^2$. However, the model can easily accommodate additional treatment conditions and covariates, and binary or count outcome variables.

@@

Obviously, such a simple, global GLM will not fit the data well, especially in the presence of moderators. For illustrational purposes, however, we take it as a starting point to illustrate MOB. The global GLM fitted to the motivating example dataset is depicted in Figure~\ref{fig:fixedeffects}. As the boxplots show, there is little difference between the global effects of the two treatments and there is considerable residual variance.

The MOB algorithm can be used to partition the dataset using additional covariates and find better fitting local models. To this end, the MOB algorithm tests for parameter stability over a set of auxiliary covariates, also known as \textit{partitioning variables}. When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor. To find these partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) statistically test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.

In step (2) a test statistic quantifying parameter instability is calculated for every potential partitioning variable. As the distribution of these test statistics under the null hypothesis of parameter stability is known, a $p$-value for every partitioning variable can be calculated. Note that a more in-depth discussion of the parameter stability tests is outside the scope of this paper, but can be found in \citeA{ZeilyHorn07} and \citeA{ZeilyHoth08}.

If at least one of the partitioning variables yields a $p$-value below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3). This partition is created using $U_{{k^*}}$, the partitioning variable with the minimal $p$-value in step (2). The split point for $U_{k^*}$ is selected by taking the value that minimizes the sum of the values of the objective function in both partitions. In other words, for every possible split point, the objective function is minimized separately in the two resulting subsets of the data; the split point yielding the minimum sum of the objective functions is selected. In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected (or the subsets become too small).

@@

To illustrate, we fitted a GLM tree on the artificial motivating dataset. In addition to the treatment indicator and treatment outcome used to fit the earlier GLM, we specified the anxiety, duration and age variables as potential partitioning variables. Figure~\ref{fig:example_mobtree} shows the resulting GLM tree. MOB partitioned the observations into four subgroups, each with a different estimate $\beta_j$. Age was correctly not identified as a partitioning variable and the left- and rightmost nodes are in accordance with the true treatment-subgroup interactions described above. However, the two nodes in the middle do not represent true subgroups, possibly due to the dependence of observations within clusters not being taken into account.

@@

\begin{figure}[tb!]
\centering
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree)
@
\caption{\small GLMM tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
\label{fig:example_glimmertree}
\end{figure}

>>>>

We will use an artificial motivating dataset with treatment-subgroup interactions to introduce the GLMM tree algorithm. The R code used to generate the example dataset is provided in the Appendix \readme{I would put this into the appendix of the vignette instead.}.

The dataset consists of a set of observations on $N = 150$ patients, who were randomly assigned to one of two treatment alternatives (Treatment~1 or Treatment~2). The treatment outcome is represented by the variable depression, quantifying post-treatment depressive symptomatology. The potential moderator variables are duration, age and anxiety. Duration reflects the number of months the patient has been suffering from depression prior to treatment, age reflects patients' age in years at the start of treatment and anxiety reflects patients' total scores on an anxiety inventory administered before treatment. Finally, each patient was part of one of ten clusters, each having a different value for the random intercept, uncorrelated with the partitioning variables.

The outcome variable was generated such that there are three subgroups with differential treatment effectiveness, corresponding to the terminal nodes in Figure~\ref{fig:exampletree}: For the first subgroup of patients (node~3) with short duration ($\le 8$) months of depressive symptoms prior to treatment and low anxiety scores ($\le 10$), Treatment~1 leads to lower post-treatment depression than in Treatment~2. For the second subgroup of patients (node~4) with short duration but high anxiety scores ($> 10$) post-treatment depression is about equal in both treatment conditions. For the third subgroup of patients (node~5) with long duration ($> 8$ months) Treatment~2 leads to lower post-treatment depression than Treatment~1. Thus, duration and anxiety are true partitioning or moderator variables, whereas age is not. Anticipating the final results of our analyses, the treatment-subgroup interactions are depicted in Figure~\ref{fig:example_glimmertree}, which shows the GLMM tree that accurately recovered the treatment-subgroup interactions. 
\readme{I've incorporated a characterization of the groups in terms of duration and anxiety here. But probably a better description of these variables is needed earlier. Possibly in a table with some summary statistics, e.g., mean/sd/range. Finally, the description in the manual Rd pages of the package should also be improved.}

@@

The rationale behind MOB is that a single global GLM (or other parametric model) may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find better-fitting models in each cell of the partition.
For example, to assess the effect of treatment, we may first fit a global GLM where the treatment indicator has the same effect/coefficient on the outcome for all observations. Subsequently, the data may be partitioned recursively with respect to other covariates, leading to separate models with different treatment effects/coefficients in each subsample.
\readme{Expanded on the treatment GLM MOB explanation to give more context and motivation.}

More formally, in a single global GLM the expectation $\mu_i$ of outcome $y_i$ given the treatment regressor $x_i$ is modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
	\label{eq:expected_value}
	E[y_i | x_i] & = & \mu_i, \\
	\label{eq:fixedeffects}
	g(\mu_{i}) & = & x_{i}^{\top}\beta,
\end{eqnarray}
%
where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function. $\beta$ is a vector of fixed-effects regression coefficients. For simplicity, in the current paper we focus on two treatment groups and no further covariates in the GLM, so that in our illustrations $x_{i}$ and $\beta$ both have length 2. For the continuous response variable in the motivating data set, we employ the identity link function and assume a normal distribution for the error (denoted by $\epsilon_i = y_i - \mu_i$) with mean zero and variance $\sigma_{\epsilon}^2$. Thus, the first element of $\beta$ then corresponds to the mean of the linear predictor in the first treatment group and the second element corresponds to the mean difference in the linear predictor between the first and second treatment groups. However, the model can easily accommodate additional treatment conditions and covariates, as well as binary or count/Poisson outcome variables.

@@

Obviously, such a simple, global GLM will not fit the data well, especially in the presence of moderators. For expository purposes, however, we take it as a starting point to illustrate MOB. The global GLM fitted to the motivating example dataset is depicted in Figure~\ref{fig:fixedeffects}. As the boxplots show, there is little difference between the global effects of the two treatments and there is considerable residual variance.

The MOB algorithm can be used to partition the dataset using additional covariates and find better-fitting local models. To this end, the MOB algorithm tests for parameter stability over a set of auxiliary covariates, also called \textit{partitioning variables}. When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor. To find these partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) statistically test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.

In step (2) a test statistic quantifying parameter instability is calculated for every potential partitioning variable. As the distribution of these test statistics under the null hypothesis of parameter stability is known, a $p$-value for every partitioning variable can be calculated. Note that a more in-depth discussion of the parameter stability tests is beyond the scope of this paper, but can be found in \citeA{ZeilyHorn07} and \citeA{ZeilyHoth08}.

If at least one of the partitioning variables yields a $p$-value below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3). This partition is created using $U_{{k^*}}$ \readme{the U variables have not been introduced, no symbol has been used for the partitioning variables so far}, the partitioning variable with the minimal $p$-value in step (2). The split point for $U_{k^*}$ is selected by taking the value that minimizes the sum of the values of the objective function in both partitions. In other words, for every possible split point, the objective function is minimized separately in the two resulting subsets of the data; the split point yielding the minimum sum of the objective functions is selected. In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected (or the subsets become too small).

@@

To illustrate, we fitted a GLM tree on the artificial motivating dataset. In addition to the treatment indicator and treatment outcome used to fit the earlier GLM, we specified the anxiety, duration and age variables as potential partitioning variables. Figure~\ref{fig:example_mobtree} shows the resulting GLM tree. MOB partitioned the observations into four subgroups, each with a different estimate $\beta_j$. Age was correctly not identified as a partitioning variable and the left- and rightmost nodes are in accordance with the true treatment-subgroup interactions described above. However, the two nodes in the middle have an unnecessary split and thus do not represent true subgroups, possibly due to the dependence of observations within clusters not being taken into account.

@@

\begin{figure}[tb!]
\centering
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree)
@
\caption{\small GLMM tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months) and age were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
\label{fig:example_glimmertree}
\end{figure}
